Optimistic Atomic Multicast in One Communication Delay

### 2011-07-19 textos/optpaxos/draft_dropbox.tex
adding some more revision control to dropbox

\newcommand{\mv}[1]{\ensuremath{\operatorname{\mathit{#1}}}}
\definecolor{dark}{gray}{.6}
\newcommand{\bc}[1]{\textcolor{dark}{#1}}
\newtheorem{lems}{Lemma}
\newtheorem{props}{Proposition}
\newtheorem{thms}{Theorem}
\newtheorem{defs}{Definition}
\newtheorem{obs}{Observation}

@@

Moreover, each process $p$ of group $G$ keeps some lists of messages:
\begin{itemize}
  %\item \textit{optPending}, which contains the messages waiting to be \mbox{\textit{\opt{}ed}} (optimistically delivered);
  \item \textit{propPending}, containing the messages waiting to be proposed by $p$;
  \item \textit{barPending}, with the messages ready to be \cons{}ed, but which may be waiting for barriers from the groups in $sendersTo(G)$;
  \item \textit{decided}, which contains the messages that have already been proposed and decided within $group(p)$;
  \item \textit{delivered}, which contains the messages that have been \cons{}ed already.
\end{itemize}

@@

  When a process in $G$ \amcast{}s $m$, it is first \rmcast{} to the other processes of $G$. Then, when $m$ is \rmdel{}ed at $p$, $p$ checks whether the latest consensus instance $I_{k}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $propPending$ queue as the next message being proposed by it in the consensus instance $I_{k+1}$, so other tasks can keep being executed. Then, once $I_{k}$ has been decided, $p$ may start a new instance. Before that, all messages in $propPending$ that have been already decided are discarded from $propPending$. The rest is proposed as a batch in $I_{k+1}$.

@@

  \end{itemize}
%   
%   In this case, when $m$ is received by $p$, $p$ checks whether the latest consensus instance $I_{k}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $propPending$ queue as the next message being proposed by it in the consensus instance $I_{k+1}$, so other tasks can keep being executed. Then, once $I_{k}$ has been decided, $p$ may start a new instance. Before that, all messages in $propPending$ that have been already decided are discarded from $propPending$. The rest is proposed as a batch in $I_{k+1}$. Once $m$ is decided, it is not immediately delivered to the application. Instead, $p$ checks whether some message $m_{prv} : m_{prv}.ts \geq m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, $m$ is inserted into a $barPending$ list for later being delivered, which will happen once every group $G'$ in $sendersTo(G)$ has already sent a message $barrier(G',G) = t$, such that $t > m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.
% 
%   \item When $m$ is originated in $G$, but it has at least one group other than $G$ as a destination:
%   
%   In this case, when $m$ is received, $p$ tries to initiate a consensus instance within $G$ to decide $m$. If $p$ cannot start the proposal now, $m$ is enqueued in $propPending$ for being proposed later along with other pending messages. Then, once $p$ may start a new instance, all messages in $propPending$ that have been already decided are discarded from $propPending$. The rest is proposed as a batch in the new consensus instance. Once any message $m$ is decided, $p$ checks whether some message $m_{prv} : m_{prv}.ts > m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, if $G \in m.dst$, $m$ is inserted in the $barPending$ list. Besides, when $m$ is decided, $p$ sends $m$ to every \mbox{$p' \in (m.dst \setminus \{G\})$}. When $m$ is received by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $barPending$ list. If not, $p'$ inserts $m$ into $barPending$ and adjusts $barrier(G, G')$ to $m.ts$. To ensure that, once a message $m$ is received from another group $G$, every message $m' : m'.ts<m.ts$ also from $G$ has already been received, every message is sent through a lossless FIFO channel\footnote{An ordinary TCP connection would be enough to provide such FIFO lossless channel. Here, we use FIFO reliable multicast.}.

@@

\blankline
\INDENT{Initialization}
  \STATE $k \leftarrow 0$, $nextProp \leftarrow 0$, $decided \leftarrow \emptyset$, $delivered \leftarrow \emptyset$, $propPending \leftarrow \emptyset$, $barPending \leftarrow \emptyset$
  \INDENT{\textbf{for all} $h \in sendersTo(g)$ \textbf{do}}
    \STATE $barrier(h) \leftarrow (0,0)$ 
  \ENDINDENT
\ENDINDENT

@@

\blankline
\WHEN{\rmdelarg{$m$}}
  \IF{$g = m.src$}
    \STATE $propPending \leftarrow propPending \cup \{m\}$ \label{algline:addtoproppending}
  \ELSIF{$m \notin barPending \wedge m \notin delivered$}
    \STATE $barPending \leftarrow barPending \cup \{m\}$ \label{algline:insbp1}
    \STATE $barrier(m.src) \leftarrow m.ts$ \label{algline:incbar}
  \ENDIF
\ENDWHEN

\blankline
\WHEN{$propPending \setminus decided \neq \emptyset$}
	\STATE $propPending \leftarrow propPending \setminus decided$
	\STATE propose($k, propPending$) \label{algline:propose}
	\STATE \textbf{wait until} decide($k, msgSet$)
	\WHILE{$msgSet \setminus decided \neq \emptyset$}
		\STATE let $m$ be the message in $msgSet \setminus decided$ with smallest timestamp
		\STATE let $m'$ be the message in $decided$ with greatest timestamp
		\IF{$m'.ts > m.ts$}
			\STATE $m.ts \leftarrow (m'.ts.rtc, m'.ts.seq + 1)$
		\ENDIF     
		\IF{$g \in m.dst$} \label{algline:checkcons}
			\STATE $barPending \leftarrow barPending \cup \{m\}$  \label{algline:insbp2}
		\ENDIF
		\STATE $decided \leftarrow decided \cup \{m\}$  \label{algline:addtodecided}
		\STATE \rmcastarg{$m.dst \setminus \{g\}$}{$m$} \label{algline:rmothers}
	\ENDWHILE
	\STATE $k \leftarrow k + 1$
\ENDWHEN

\blankline
%\WHEN{$\exists m \in barPending : \forall G' \in sendersTo(G): m.ts < barrier(G',G)$\\ $\wedge\ \nexists m' \in barPending : m'.ts < m.ts$} \label{algline:inbarpending}
\WHEN{$barPending \setminus delivered \neq \emptyset$} \label{algline:inbarpending}
	\STATE let $m$ be the message in $barPending \setminus delivered$ with smallest timestamp
	\IF{$\forall h \in sendersTo(g): m.ts < barrier(h)$}
%		\STATE $barPending \leftarrow barPending \setminus \{m\}$ \label{algline:rmfrombp}
		\IF {$m \neq null$}
		  \STATE \amdelarg{$m$} \label{algline:consdeliver}
		\ENDIF
		\STATE $delivered \leftarrow delivered \cup \{m\}$ \label{algline:addtodelivered}
	\ENDIF
\ENDWHEN
\blankline

@@

\blankline
\INDENT{\textit{When $\exists h: g \in sendersTo(h) \wedge getTime() - lastBarrierCreated(h) > barrierThreshold$}}
  \STATE $null \leftarrow$ empty message
  \STATE $null.ts \leftarrow (lastBarrierCreated(h) + barrierThreshold, 0)$ \label{algline:samenullid}
  \STATE $null.src \leftarrow g$
  \STATE $null.dst \leftarrow \{h\}$ 
  \STATE $propPending \leftarrow propPending \cup \{null\}$
  \COMMENT{saving that nothing was sent until null.ts}
\ENDINDENT

@@

It is necessary to guarantee that a $null$ message will eventually be proposed, decided, and a barrier will be sent to some group which might be needing it, so that progression is guaranteed. Therefore, this kind of messages are created by every process in the group, since if any one of them does not have it, such message might never be decided. %Besides, they are excluded from the $propPending$ list only when decided in the group (l. \ref{algline:nullstays} of Algorithm \ref{algorithm:deliveryminimal}). Since they are obviously never delivered to the application, they can remain in such list and be decided even after some message with higher timestamp is decided. Not removing them unless they are decided ensures that they will be sent and that other groups will be able to increase their barrier values. % On the other hand, this brings another problem: if every process in $G$ creates a different $null$ message, many $null$ messages might have to be decided before a meaningful message.
To prevent the multiple $null$ messages -- created by different processes within the group -- of being decided, they could be created in a way such that the different processes can somehow figure out that two different $null$ messages are equivalent\footnote{This could be done by assuming no timestamp collisions and by using them to uniquely identify messages. Then, only one of the messages created with the same timestamp (l. \ref{algline:samenullid} of Algorithm \ref{algorithm:nullperiodic}) would be decided.}.

@@

When the process $p'$ of some group $G'$ receives a barrier request for a message $m$, %such that $G' \in blockers(m)$,
$p'$ knows that there are other groups depending on the barrier of $G'$ to deliver $m$. For that reason, it will immediately create a $null$ message with a timestamp equal to $m.ts$ and with $null.dst = m.dst$ $\cap$ $receiversFrom(G')$. Then, $p'$ will insert such message in its %$optPending$
$propPending$ queue. % -- to ensure that, once it goes to $propPending$, any message \mbox{$m': m'.ts < null.ts \wedge m'.src = G'$} is already there. As soon as $null.ts > now - w(p')$, the 
Once the $null$ message is proposed and decided in $G'$, each process of $G'$ will send it to every process in each group \mbox{$G \in m.dst \cap receiversFrom(G')$}. This way, any group which was waiting for a barrier from $G'$ to deliver $m$ will be able to do so as soon as it receives such $null$ message. The new delivery algorithm would be as described in \mbox{Algorithm {\ref{algorithm:nullondemand}}}.

@@

\blankline
\INDENT {Initialization}
  \STATE $k \leftarrow 0$, $nextProp \leftarrow 0$, $decided \leftarrow \emptyset$, $delivered \leftarrow \emptyset$, $propPending \leftarrow \emptyset$, $barPending \leftarrow \emptyset$
  \INDENT{\textbf{for all} $G' \in sendersTo(G)$ \textbf{do}}
    \STATE $barrier(G',G) \leftarrow -\infty$ 
  \ENDINDENT
\ENDINDENT

@@

\blankline
\INDENT{\textit{When} \rmdelarg{$m'$}}
  \IF{$G = m'.src$}
    \STATE $propPending \leftarrow propPending \cup \{m'\}$
  \ELSIF{$m' \notin barPending \wedge m' \notin delivered$}
    \STATE $barrier(m'.src,G) \leftarrow m'.ts$ \label{algline:incbar}
    \IF {$G \in m'.dst$}
      \STATE $barPending \leftarrow barPending \cup \{m'\}$
    \ENDIF
    \IF{$G \in blockers(m') \wedge m' \neq null $}
      \STATE $null \leftarrow$ empty message
      \STATE $null.src \leftarrow G$
      \STATE $null.ts \leftarrow m'.ts$ \label{algline:nulltsmts}
      \STATE $null.dst \leftarrow m'.dst \cap receiversFrom(G)$
      \STATE $propPending \leftarrow propPending \cup \{null\}$
    \ENDIF
  \ENDIF
\ENDINDENT

\blankline
\INDENT{\textit{When $\exists m \in propPending \wedge nextProp = k$}}
%  \STATE $propPending \leftarrow propPending \setminus \{m\}$
%  \IF {$m \notin decided \wedge \nexists m' \in decided: m'.ts > m.ts$}
    \STATE $propPending \leftarrow propPending \setminus decided$%(decided \cup \{m' : \exists m'' \in decided \wedge m'.ts < m''.ts \wedge m' \neq null\})$ 
    \label{algline:nullstays}
    \IF {$propPending \neq \emptyset$}
      \STATE $nextProp \leftarrow k + 1$
      \STATE Propose($k$,$propPending$)
    \ENDIF
%  \ENDIF
\ENDINDENT

@@

 Assuming $\delta$ as the communication delay between every pair of processes, the time needed to decide a message $m$ after it has been \amcast{} by some process is equal, in the worst case, to $\delta + 2\tconsm$, where $\delta$ is the time needed to \rmdel{} a message and \tcons\ is the time needed to execute a consensus instance. This value is counted twice because $m$ may have been inserted into $propPending$ right after a consensus instance has been initiated. In that case, $m$ would have to wait such consensus to finish, to be then proposed and finally decided. However, it may also happen that $m$ has been inserted into $propPending$ right before some proposal has been made, so it would take only $\delta + \tconsm$ to decide $m$. As $m$ may go into $propPending$ anytime between the worst and the best case with the same probability, the average time needed for deciding $m$ would be equal to $\delta + 1.5\tconsm$. Nevertheless, the time needed to finally \cons{} $m$ will depend on when barriers are received from other groups.

@@

%   \STATE $k \leftarrow 0$, $nextProp \leftarrow 0$, $decided \leftarrow \emptyset$, $propPending \leftarrow \emptyset$, $optPending \leftarrow \emptyset$, $barPending \leftarrow \emptyset$

@@

%     \STATE $propPending \leftarrow propPending \cup \{m\}$

@@

% \INDENT{\textit{When $\exists m \in propPending \wedge nextProp = k$}}
% %  \STATE $propPending \leftarrow propPending \setminus \{m\}$

@@

%     \STATE $propPending \leftarrow propPending \setminus (decided \cup \{m' : \exists m'' \in decided \wedge m'.ts < m''.ts \wedge m' \neq null\})$ \label{algline:keepnull}
%     \IF {$propPending \neq \emptyset$}

@@

%       \STATE Propose($k$,$propPending$)

@@

For a message $m$ to be delivered (l. \ref{algline:consdeliver}), it must be contained in the $barPending$ list (from the condition in line \ref{algline:inbarpending}).  There are two possibilities to when $m$ has been inserted into such list:
\begin{itemize}
  \item $m$ has been originated in the same same group of $p$, that is, $p \in m.dst$, which means $m$ was inserted into $barPending$ in line \ref{algline:insbp2} of Algorithm \ref{algorithm:deliveryminimal}, or
  \item $m$ has been sent from a group $G \neq G'$, which means that it was inserted into $barPending$ in line \ref{algline:insbp1}. 
\end{itemize}

In line \ref{algline:insbp2}, $m$ is inserted into $barPending$ only if $G \in m.dst$, which satisfies condition (2), and if it was decided in some consensus instance within $G$. To have been decided within $G$, from the properties of consensus \cite{lamport1998ptp}, we know that it must have been proposed by some process of $G$, which happens in line \ref{algline:propose}. In line \ref{algline:propose}, for a message to be proposed, it must be in the $propPending$, as its contents are the value proposed. For a message to be in the $propPending$ list, it must have been inserted there, which happens only in line \ref{algline:addtoproppending}. For l. \ref{algline:addtoproppending} to be executed, $m$ must have been \rmdel{}ed and $G$ must be the source group of $m$, that is, $m.src = G$. For a message to be \rmcast{} by a process to its own group, a \amcastarg{$m$} call must have been made -- which satisfies condition (1) --, for line \ref{algline:rmlocal} is the only one where a process \rmcast{}s a message to its own group.

Also, for a message to be inserted into $propPending$ (l. \ref{algline:addtoproppending}) it cannot have been decided already (l. \ref{algline:removedecided}). We assume here that a process can only make a proposal for consensus instance $k$ when it knows all the values decided in all instances prior to $k$. Therefore, when a process proposes a message, it knows all the messages which were already decided in previous consensus instances. As a decided message is inserted into a $decided$ list (l. \ref{algline:addtodecided}, which is always executed after a decision), a process cannot propose a message which was decided already and, therefore, a message cannot be decided twice (in two different consensus instances). As a message is inserted into $barPending$ by processes of its own group of origin only when decided (l. \ref{algline:insbp2}), it cannot be inserted twice, thus it cannot be delivered twice, satisfying condition (3).

>>>>

\newcommand{\mv} [1]{\ensuremath{\operatorname{\mathit{#1}}}}
\definecolor{dark}{gray}{.6}
\newcommand{\bc}[1]{\textcolor{dark}{#1}}
\newtheorem{lems}{Lemma}
\newtheorem{props}{Proposition}
\newtheorem{thms}{Theorem}
\newtheorem{defs}{Definition}
\newtheorem{obs}{Observation}

@@

Moreover, each process $p$ of group $G$ keeps some lists of messages:
\begin{itemize}
  %\item \textit{optPending}, which contains the messages waiting to be \mbox{\textit{\opt{}ed}} (optimistically delivered);
  \item \textit{groupMsgs}, containing the messages waiting to be proposed by $p$;
  \item \textit{barPending}, with the messages ready to be \cons{}ed, but which may be waiting for barriers from the groups in $sendersTo(G)$;
  \item \textit{decided}, which contains the messages that have already been proposed and decided within $group(p)$;
  \item \textit{delivered}, which contains the messages that have been \cons{}ed already.
\end{itemize}

@@

  When a process in $G$ \amcast{}s $m$, it is first \rmcast{} to the other processes of $G$. Then, when $m$ is \rmdel{}ed at $p$, $p$ checks whether the latest consensus instance $I_{k}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $groupMsgs$ queue as the next message being proposed by it in the consensus instance $I_{k+1}$, so other tasks can keep being executed. Then, once $I_{k}$ has been decided, $p$ may start a new instance. Before that, all messages in $groupMsgs$ that have been already decided are discarded from $groupMsgs$. The rest is proposed as a batch in $I_{k+1}$.

@@

  \end{itemize}
%   
%   In this case, when $m$ is received by $p$, $p$ checks whether the latest consensus instance $I_{k}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $groupMsgs$ queue as the next message being proposed by it in the consensus instance $I_{k+1}$, so other tasks can keep being executed. Then, once $I_{k}$ has been decided, $p$ may start a new instance. Before that, all messages in $groupMsgs$ that have been already decided are discarded from $groupMsgs$. The rest is proposed as a batch in $I_{k+1}$. Once $m$ is decided, it is not immediately delivered to the application. Instead, $p$ checks whether some message $m_{prv} : m_{prv}.ts \geq m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, $m$ is inserted into a $barPending$ list for later being delivered, which will happen once every group $G'$ in $sendersTo(G)$ has already sent a message $barrier(G',G) = t$, such that $t > m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.
% 
%   \item When $m$ is originated in $G$, but it has at least one group other than $G$ as a destination:
%   
%   In this case, when $m$ is received, $p$ tries to initiate a consensus instance within $G$ to decide $m$. If $p$ cannot start the proposal now, $m$ is enqueued in $groupMsgs$ for being proposed later along with other pending messages. Then, once $p$ may start a new instance, all messages in $groupMsgs$ that have been already decided are discarded from $groupMsgs$. The rest is proposed as a batch in the new consensus instance. Once any message $m$ is decided, $p$ checks whether some message $m_{prv} : m_{prv}.ts > m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, if $G \in m.dst$, $m$ is inserted in the $barPending$ list. Besides, when $m$ is decided, $p$ sends $m$ to every \mbox{$p' \in (m.dst \setminus \{G\})$}. When $m$ is received by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $barPending$ list. If not, $p'$ inserts $m$ into $barPending$ and adjusts $barrier(G, G')$ to $m.ts$. To ensure that, once a message $m$ is received from another group $G$, every message $m' : m'.ts<m.ts$ also from $G$ has already been received, every message is sent through a lossless FIFO channel\footnote{An ordinary TCP connection would be enough to provide such FIFO lossless channel. Here, we use FIFO reliable multicast.}.

@@

\blankline
\INDENT{Initialization}
  \STATE $k \leftarrow 0$, $nextProp \leftarrow 0$, $decided \leftarrow \emptyset$, $delivered \leftarrow \emptyset$, $groupMsgs \leftarrow \emptyset$, $barPending \leftarrow \emptyset$
  \INDENT{\textbf{for all} $h \in sendersTo(g)$ \textbf{do}}
    \STATE $barrier(h) \leftarrow (0,0)$ 
  \ENDINDENT
\ENDINDENT

@@

\blankline
\WHEN{\rmdelarg{$m$}}
  \IF{$g = m.src$}
    \STATE $groupMsgs \leftarrow groupMsgs \cup \{m\}$ \label{algline:addtogroupMsgs}
  \ELSIF{$m \notin barPending \wedge m \notin delivered$}
    \STATE $barPending \leftarrow barPending \cup \{m\}$ \label{algline:insbp1}
    \STATE $barrier(m.src) \leftarrow m.ts$ \label{algline:incbar}
  \ENDIF
\ENDWHEN

\blankline
\WHEN{$groupMsgs \setminus decided \neq \emptyset$}
  \STATE $k \leftarrow k + 1$
	\STATE $undecided \leftarrow prop \setminus decided$
	\STATE propose($k, undecided$) \label{algline:propose}
	\STATE \textbf{wait until} decide($k, msgSet$)
  \STATE $msgSet \leftarrow msgSet \setminus decided$
	\WHILE{$msgSet \setminus decided \neq \emptyset$}
		\STATE let $m$ be the message in $msgSet \setminus decided$ with smallest timestamp
		\STATE let $m'$ be the message in $decided$ with greatest timestamp
		\IF{$m'.ts > m.ts$}
			\STATE $m.ts \leftarrow (m'.ts.rtc, m'.ts.seq + 1)$
		\ENDIF     
		\IF{$g \in m.dst$} \label{algline:checkcons}
			\STATE $barPending \leftarrow barPending \cup \{m\}$  \label{algline:insbp2}
		\ENDIF
		\STATE $decided \leftarrow decided \cup \{m\}$  \label{algline:addtodecided}
		\STATE \rmcastarg{$m.dst \setminus \{g\}$}{$m$} \label{algline:rmothers}
	\ENDWHILE

\ENDWHEN

\blankline
%\WHEN{$\exists m \in barPending : \forall G' \in sendersTo(G): m.ts < barrier(G',G)$\\ $\wedge\ \nexists m' \in barPending : m'.ts < m.ts$} \label{algline:inbarPending}
\WHEN{$barPending \setminus delivered \neq \emptyset$} \label{algline:inbarPending}
	\STATE let $m$ be the message in $barPending \setminus delivered$ with smallest timestamp
	\IF{$\forall h \in sendersTo(g): m.ts < barrier(h)$}
%		\STATE $barPending \leftarrow barPending \setminus \{m\}$ \label{algline:rmfrombp}
		\IF {$m \neq null$}
		  \STATE \amdelarg{$m$} \label{algline:consdeliver}
		\ENDIF
		\STATE $delivered \leftarrow delivered \cup \{m\}$ \label{algline:addtodelivered}
	\ENDIF
\ENDWHEN
\blankline

@@

\blankline
\INDENT{\textit{When $\exists h: g \in sendersTo(h) \wedge getTime() - lastBarrierCreated(h) > barrierThreshold$}}
  \STATE $null \leftarrow$ empty message
  \STATE $null.ts \leftarrow (lastBarrierCreated(h) + barrierThreshold, 0)$ \label{algline:samenullid}
  \STATE $null.src \leftarrow g$
  \STATE $null.dst \leftarrow \{h\}$ 
  \STATE $groupMsgs \leftarrow groupMsgs \cup \{null\}$
  \COMMENT{saving that nothing was sent until null.ts}
\ENDINDENT

@@

It is necessary to guarantee that a $null$ message will eventually be proposed, decided, and a barrier will be sent to some group which might be needing it, so that progression is guaranteed. Therefore, this kind of messages are created by every process in the group, since if any one of them does not have it, such message might never be decided. %Besides, they are excluded from the $groupMsgs$ list only when decided in the group (l. \ref{algline:nullstays} of Algorithm \ref{algorithm:deliveryminimal}). Since they are obviously never delivered to the application, they can remain in such list and be decided even after some message with higher timestamp is decided. Not removing them unless they are decided ensures that they will be sent and that other groups will be able to increase their barrier values. % On the other hand, this brings another problem: if every process in $G$ creates a different $null$ message, many $null$ messages might have to be decided before a meaningful message.
To prevent the multiple $null$ messages -- created by different processes within the group -- of being decided, they could be created in a way such that the different processes can somehow figure out that two different $null$ messages are equivalent\footnote{This could be done by assuming no timestamp collisions and by using them to uniquely identify messages. Then, only one of the messages created with the same timestamp (l. \ref{algline:samenullid} of Algorithm \ref{algorithm:nullperiodic}) would be decided.}.

@@

When the process $p'$ of some group $G'$ receives a barrier request for a message $m$, %such that $G' \in blockers(m)$,
$p'$ knows that there are other groups depending on the barrier of $G'$ to deliver $m$. For that reason, it will immediately create a $null$ message with a timestamp equal to $m.ts$ and with $null.dst = m.dst$ $\cap$ $receiversFrom(G')$. Then, $p'$ will insert such message in its %$optPending$
$groupMsgs$ queue. % -- to ensure that, once it goes to $groupMsgs$, any message \mbox{$m': m'.ts < null.ts \wedge m'.src = G'$} is already there. As soon as $null.ts > now - w(p')$, the 
Once the $null$ message is proposed and decided in $G'$, each process of $G'$ will send it to every process in each group \mbox{$G \in m.dst \cap receiversFrom(G')$}. This way, any group which was waiting for a barrier from $G'$ to deliver $m$ will be able to do so as soon as it receives such $null$ message. The new delivery algorithm would be as described in \mbox{Algorithm {\ref{algorithm:nullondemand}}}.

@@

\blankline
\INDENT {Initialization}
  \STATE $k \leftarrow 0$, $nextProp \leftarrow 0$, $decided \leftarrow \emptyset$, $delivered \leftarrow \emptyset$, $groupMsgs \leftarrow \emptyset$, $barPending \leftarrow \emptyset$
  \INDENT{\textbf{for all} $G' \in sendersTo(G)$ \textbf{do}}
    \STATE $barrier(G',G) \leftarrow -\infty$ 
  \ENDINDENT
\ENDINDENT

@@

\blankline
\INDENT{\textit{When} \rmdelarg{$m'$}}
  \IF{$G = m'.src$}
    \STATE $groupMsgs \leftarrow groupMsgs \cup \{m'\}$
  \ELSIF{$m' \notin barPending \wedge m' \notin delivered$}
    \STATE $barrier(m'.src,G) \leftarrow m'.ts$ \label{algline:incbar}
    \IF {$G \in m'.dst$}
      \STATE $barPending \leftarrow barPending \cup \{m'\}$
    \ENDIF
    \IF{$G \in blockers(m') \wedge m' \neq null $}
      \STATE $null \leftarrow$ empty message
      \STATE $null.src \leftarrow G$
      \STATE $null.ts \leftarrow m'.ts$ \label{algline:nulltsmts}
      \STATE $null.dst \leftarrow m'.dst \cap receiversFrom(G)$
      \STATE $groupMsgs \leftarrow groupMsgs \cup \{null\}$
    \ENDIF
  \ENDIF
\ENDINDENT

\blankline
\INDENT{\textit{When $\exists m \in groupMsgs \wedge nextProp = k$}}
%  \STATE $groupMsgs \leftarrow groupMsgs \setminus \{m\}$
%  \IF {$m \notin decided \wedge \nexists m' \in decided: m'.ts > m.ts$}
    \STATE $groupMsgs \leftarrow groupMsgs \setminus decided$%(decided \cup \{m' : \exists m'' \in decided \wedge m'.ts < m''.ts \wedge m' \neq null\})$ 
    \label{algline:nullstays}
    \IF {$groupMsgs \neq \emptyset$}
      \STATE $nextProp \leftarrow k + 1$
      \STATE Propose($k$,$groupMsgs$)
    \ENDIF
%  \ENDIF
\ENDINDENT

@@

 Assuming $\delta$ as the communication delay between every pair of processes, the time needed to decide a message $m$ after it has been \amcast{} by some process is equal, in the worst case, to $\delta + 2\tconsm$, where $\delta$ is the time needed to \rmdel{} a message and \tcons\ is the time needed to execute a consensus instance. This value is counted twice because $m$ may have been inserted into $groupMsgs$ right after a consensus instance has been initiated. In that case, $m$ would have to wait such consensus to finish, to be then proposed and finally decided. However, it may also happen that $m$ has been inserted into $groupMsgs$ right before some proposal has been made, so it would take only $\delta + \tconsm$ to decide $m$. As $m$ may go into $groupMsgs$ anytime between the worst and the best case with the same probability, the average time needed for deciding $m$ would be equal to $\delta + 1.5\tconsm$. Nevertheless, the time needed to finally \cons{} $m$ will depend on when barriers are received from other groups.

@@

%   \STATE $k \leftarrow 0$, $nextProp \leftarrow 0$, $decided \leftarrow \emptyset$, $groupMsgs \leftarrow \emptyset$, $optPending \leftarrow \emptyset$, $barPending \leftarrow \emptyset$

@@

%     \STATE $groupMsgs \leftarrow groupMsgs \cup \{m\}$

@@

% \INDENT{\textit{When $\exists m \in groupMsgs \wedge nextProp = k$}}
% %  \STATE $groupMsgs \leftarrow groupMsgs \setminus \{m\}$

@@

%     \STATE $groupMsgs \leftarrow groupMsgs \setminus (decided \cup \{m' : \exists m'' \in decided \wedge m'.ts < m''.ts \wedge m' \neq null\})$ \label{algline:keepnull}
%     \IF {$groupMsgs \neq \emptyset$}

@@

%       \STATE Propose($k$,$groupMsgs$)

@@

For a message $m$ to be delivered (l. \ref{algline:consdeliver}), it must be contained in the $barPending$ list (from the condition in line \ref{algline:inbarPending}).  There are two possibilities to when $m$ has been inserted into such list:
\begin{itemize}
  \item $m$ has been originated in the same same group of $p$, that is, $p \in m.dst$, which means $m$ was inserted into $barPending$ in line \ref{algline:insbp2} of Algorithm \ref{algorithm:deliveryminimal}, or
  \item $m$ has been sent from a group $G \neq G'$, which means that it was inserted into $barPending$ in line \ref{algline:insbp1}. 
\end{itemize}

In line \ref{algline:insbp2}, $m$ is inserted into $barPending$ only if $G \in m.dst$, which satisfies condition (2), and if it was decided in some consensus instance within $G$. To have been decided within $G$, from the properties of consensus \cite{lamport1998ptp}, we know that it must have been proposed by some process of $G$, which happens in line \ref{algline:propose}. In line \ref{algline:propose}, for a message to be proposed, it must be in the $groupMsgs$, as its contents are the value proposed. For a message to be in the $groupMsgs$ list, it must have been inserted there, which happens only in line \ref{algline:addtogroupMsgs}. For l. \ref{algline:addtogroupMsgs} to be executed, $m$ must have been \rmdel{}ed and $G$ must be the source group of $m$, that is, $m.src = G$. For a message to be \rmcast{} by a process to its own group, a \amcastarg{$m$} call must have been made -- which satisfies condition (1) --, for line \ref{algline:rmlocal} is the only one where a process \rmcast{}s a message to its own group.

Also, for a message to be inserted into $groupMsgs$ (l. \ref{algline:addtogroupMsgs}) it cannot have been decided already (l. \ref{algline:removedecided}). We assume here that a process can only make a proposal for consensus instance $k$ when it knows all the values decided in all instances prior to $k$. Therefore, when a process proposes a message, it knows all the messages which were already decided in previous consensus instances. As a decided message is inserted into a $decided$ list (l. \ref{algline:addtodecided}, which is always executed after a decision), a process cannot propose a message which was decided already and, therefore, a message cannot be decided twice (in two different consensus instances). As a message is inserted into $barPending$ by processes of its own group of origin only when decided (l. \ref{algline:insbp2}), it cannot be inserted twice, thus it cannot be delivered twice, satisfying condition (3).
